A thermodynamic view of dusty protoplanetary disks

### 2017-08-08 model.tex
final revision before submission, spell-checked

In general the relative velocity $\bm{v}_\mathrm{d} -
\bm{v}_\mathrm{g}$ obeys a complicated evolutionary equation 
\citep[see, e.g.][]{youdin05a}. % obtained from taking the difference
%between evolutionary equations for $\bm{v}_\mathrm{g}$ and
%$\bm{v}_\mathrm{d}$. 
 However, for tightly-coupled dust
particles with $\tstop\OmK\ll 1 $, where $\OmK$ is the Keplerian orbital
frequency (since we are interested in protoplanetary disks), we can use the 
`terminal velocity approximation' to set %The dust-gas velocity difference 
%Then so that the dust-gas velocity
%difference is 
\begin{align}
  \bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} = \frac{\nabla
    P}{\rhog}\tstop 
\end{align}
\citep{jacquet11}. This equation reflects the well-known effect of 
particle drift towards pressure maxima \citep{weidenschilling77}.

@@

Under the terminal velocity approximation the dust-gas mixture obeys   
the first-order (in $\tstop$) one-fluid equations: 
\begin{align} 
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
   &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
     \tstop \nabla P \right),\label{dusteq}\\
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi_\mathrm{tot} - \frac{1}{\rho}\nabla  P, \label{momeq}\\ 
  &\frac{D T}{D t} = - \left(\gamma-1\right)T\nabla\cdot\bm{v} +
  \mathcal{H} + \mathcal{H}_\mathrm{eff}  - \Lambda  \label{tempeq} 
\end{align} 
\citep[see ][for a detailed derivation from the two-fluid
equations]{laibe14,price15}, where $D/Dt \equiv \p_t +
\bm{v}\cdot\nabla$ is the Lagragian 
derivative \emph{following the mixture}'s center-of-mass velocity $\bm{v}$. 
For an ideal gas the pressure is given 
by $P = \mathcal{R}\rhog T/\mu $, where $\mathcal{R}$ is
the gas constant and $\mu$ is the mean molecular weight.

@@

For the mixture's temperature evolution, Eq. \ref{tempeq}, reflects the gas energy equation. 
Here $\gamma$ is 
the adiabatic index; $\mathcal{H}$ represents gas heating and $\mathcal{H}_\mathrm{eff}$ 
is an effective source term arising from transforming from the 
two-fluid to one-fluid variables \citep[see][ for 
details]{laibe14}. We also include a simple model of radiative
cooling, 
\begin{align}
  \Lambda = \frac{T -
    T_\mathrm{ref}}{t_\mathrm{cool}}, \label{cooling} 
\end{align}
which relaxes the mixture back to a prescribed temperature profile   
$T_\mathrm{ref}$ on a timescale of $\tcool$. We will shortly simplify
the problem by considering rapid cooling, $\tcool\to0$.

@@

\subsection{Locally isothermal equation of state}\label{loc_iso_eos}
We simplify the energy equation by considering short cooling 
times $\tcool\to 0$, appropriate for the outer parts of an irradiated
protoplanetary disk \citep{chiang97,lin15}. Then the disk temperature
$T = T_\mathrm{ref}$ at all times, and so we may
adopt a locally isothermal equation of state 
\begin{align}\label{eos}
  P = c_s^2(r,z)\rhog = c_s^{2}\left(1 - \tepsilon\right)\rho,   
\end{align}
where $c_s(r,z)= \sqrt{\mathcal{R}T_\mathrm{ref}/\mu}$ is a prescribed
sound-speed profile fixed in time. In most applications we consider vertically 
isothermal disks with \begin{align}\label{power_temp}
  c_s^2(r) \propto r^{q},
\end{align}
where $q$ is the power-law index for the disk temperature. For $q=0$
the disk is strictly isothermal.

Notice Eq. \ref{eos} resembles an ideal gas equation of state but with  
a reduced temperature $\widetilde{T} = 
T_\mathrm{ref}\left(1-\tepsilon\right)$. %sound-speed $c_{s,\mathrm{eff}} = c_s\sqrt{\left(1 -
This reduced temperature decreases with dust-loading. 
%   \tepsilon\right)}$. 
%then the equation of state has the standard
%form $P=c_{s,\mathrm{eff}}^2\rho$. 
Since $\tepsilon$ typically decrease away from the midplane, we expect
vertically isothermal dusty disks to behave as if the temperture
\emph{increased} away from $z=0$.

\subsection{Effective energy equation}\label{energy_analogy}
Although we have deleted the true energy
equation (Eq. \ref{tempeq}) by fixing a locally isothermal
equation of state (Eq. \ref{eos}), we show that the mixture 
nevertheless obeys an effective energy equation. This is
because advection of the dust-fraction, described by
Eq. \ref{dusteq}, can be transformed into an energy-like 
equation. 
%We show that for a prescribed temperature distribution, the mixture
%obeys an evolutionary energy equation, due to the advection of the
%dust-fraction. 
Writing $\tepsilon=\tepsilon(P,\rho)$ using Eq. \ref{eos}
\begin{align*}
  \tepsilon = 1 - \frac{P}{c_s^2(r,z)\rho}, 
\end{align*}
then Eq. \ref{dusteq} becomes
\begin{align}
%\frac{DP}{Dt} 
\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  
&= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 -
  \frac{P}{c_s^2\rho}\right)\nabla 
  P\right].
\end{align} 
For comparison with the standard energy equation in hydrodynamics,
re-writing Eq. \ref{tempeq} for a pure ideal gas with $P \propto
\rhog T$, and reverting back to gas velocities, gives  
%For later comparison it is convenient to re-write the temperature
%evolution, Eq. \ref{tempeq}, to that for the pressure: 
\begin{align}
  \frac{\p P}{\p t} + \bm{v}_\mathrm{g}\cdot\nabla P = - \gamma P
  \nabla\cdot \bm{v}_\mathrm{g}  +
  \frac{P}{T}\left(\mathcal{H} -
    \Lambda\right).\label{real_pressure} 
%\frac{\rho \mathcal{R}}{\mu}\frac{T
%  - T_\mathrm{ref}}{t_\mathrm{cool}}, \label{realenergy}
\end{align} 
We can thus interpret Eq. \ref{eff_energy} as the energy equation for
and ideal gas with adiabatic index $\gamma=1$, but now with the
imposed temperature gradient $\nabla c_s^2$ and dust-gas drag $\mathcal{C}$ 
acting as source terms.

@@

The specific entropy of an ideal gas is 
$S = C_P\ln{\left(P^{1/\gamma}/\rhog\right)}$, where $C_P$ is the heat 
capacity at constant pressure.

Since we have shown that a locally isothermal dusty gas effectively
has $\gamma=1$ we can define an effective 
entropy for the \emph{mixture} as 
\begin{align}
   \seff \equiv \ln \frac{P}{\rho} = \ln{\left[c_s^2(1-\tepsilon)\right]}.  
\end{align} 
The constant heat capacity has been absorbed into $\seff$. 
%This follows naturally by combining the effective 
%energy Eq. \ref{eff_energy} and Eq. \ref{masseq}. 
Then 
\begin{align*}
  \frac{D \seff}{D t} = \bm{v}\cdot\nabla\ln{c_s^2} +
  \frac{c_s^2}{P}\nabla\cdot\left(\tstop\tepsilon\nabla P\right), 
\end{align*}
which is equivalent to entropy evolution in standard  
hydrodynamics, albeit with source terms. With an effective
entropy defined this way, many of the  
results concerning the stability of (locally) isothermal dusty gas
will have identical form and interpretations to that for pure gas.

@@

%\subsection{Dusty buoyancy forces}
We can now define the vertical buoyancy frequency $N_z$ of the mixture
as    
%$N_z$ of the mixture is given by 
\begin{align}
  N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p \seff}{\p z}
  =c_s^2(r) \frac{\p \ln{\rhog}}{\p z}\frac{\p \tepsilon}{\p z},
%  \quad 
 % N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p \seff}{\p z}.
%&=
  %\frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\frac{\p\ln\rhog}{\p 
%  z}\frac{\p\epsilon}{\p z} \\ &
  %                                =
 % \frac{\epsilon}{\left(1+\epsilon\right)^2}\left(\frac{z}{\Htilde}\right)^2\OmK^2\notag,  
\end{align}
%Typically in a thin, smooth disk we have $|N_r|\ll |N_z|$. 
where the second equality assumes a vertically isothermal disk. 
Protoplanetary disks have $\p_z\rhog<0$, thus stability against vertical 
convection ($N_z^2>0$) requires $\p_z\tepsilon <0 $, i.e. the dust density drops faster 
than the gas density away from the midplane. 
This is equivalent to 
entropy increasing away from the midplane. A similar expression exists
for the radial buoyancy frequency $N_r$, but $|N_r|\ll |N_z|$ in thin,
smooth disks.

@@

%{\bf
\subsection{Physical disk conditions}
%physical disk conditions underwhich the above approx
%(small/well-coupled dust and isothermal gas) is applicable. smaller
%than mm at few tens of au. give dimensional numbers here
%}
The above correspondence between isothermal dusty gas and adiabatic
pure gas is derived under the terminal velocity approximation, which applies
to short stopping times, $\tstop\OmK\ll 1$; and the locally isothermal
approximation, which applies to short cooling times, $t_\mathrm{cool}\OmK\ll
1$. In typical protoplanetary disk
models such as the Minimum Mass Solar Nebula, $\tstop$ decreases for
smaller particles and/or at smaller radii \citep{youdin11}. However,
$t_\mathrm{cool}$ is generally only small in the outer disk
\citep{lin15,malygin17}. Combining these results, we estimate the 
correspondence will be applicable to particles less than mm in size at a few
to 10s of AU in protoplanetary disks. However, note that the
isothermal approximation may be relaxed to other fixed equations of
state (see \S\ref{gen_poly}).

@@

For a given distribution of the dust-fraction $\tepsilon$ (or
dust-to-gas ratio $\epsilon$), the 
mass and momentum Eqs. \ref{masseq}---\ref{momeq} admit     
solutions with $\rho(r,z)$ and 
$\bm{v}=r\Omega(r,z)\hat{\bm{\phi}}$ where $\Omega = v_\phi/r$, which satisfy 
\begin{align}
  r\Omega^2 &= \frac{\p \Phi}{\p r} + \frac{1}{\rho}\frac{\p P}{\p
    r},\label{steady_momr}\\
  0 & = \frac{\p\Phi}{\p z} + \frac{1}{\rho}\frac{\p P}{\p z},\label{steady_momz}
%  0 & = \nabla\cdot\left(\tepsilon\tstop\nabla P\right) \label{steady_dust}
\end{align}
with $P=P(\tepsilon,\rho)$ given by the equation of state
(Eq. \ref{eos}). An explicit solution is presented in 
\S\ref{steady_state}, when we analyse the stability of
protoplanetary disks.

@@

Our equilibrium solutions satisfy the hydrostatic constraints of
Eq. \ref{steady_momr}---\ref{steady_momz} but are not in general
steady state solutions to the energy 
equation (\ref{eff_energy}), because the source term $\mathcal{C} \ne0$ for realistic disks.  
Limiting cases with $\mathcal{C} \equiv 0$ include \begin{inparaenum}[1)] 
\item 
  unstratified or 2D, razor-thin disk models with finite dust-gas drag; %(\S\ref{si});   
\item 
  perfectly-coupled dust with $\tstop=0$. %(\S\ref{results}). 
\end{inparaenum} 
The following analyses thus apply strictly to these limiting cases of
$\mathcal{C} = 0$ in equilibrium.  Our analysis is also a good approximation when 
$\tstop$ is sufficiently small such that the evolution of the background disk (e.g. dust-settling timescales) 
occur on much longer timescales than the dust dynamics of interest (e.g. instability growth timescales).

@@

In Appendix \ref{var_prin} we derive linearize Eqs. \ref{masseq}, \ref{momeq}, and 
Eq. \ref{eff_energy} to derive the following integral relation: 
\begin{align}
%&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
  \sigma^2\mathcal{I}^2
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \left(\dd v_z \dd v_r^* + \dd v_z^*\dd v_r\right) B +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\Bigl\lvert \nabla\cdot\left(P\dd
  \bm{v}\right)\Bigr\rvert^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV,\label{int_rel}
\end{align} 
%The real integral $\mathcal{I}^2>0$ 
where $^*$ denotes the complex conjugate, 
\begin{align}
  \mathcal{I}^2 \equiv \int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV
\end{align}
is the meridional kinetic energy, 
and coefficients $A,B,D$ can be 
read off Eq. \ref{integral_ex}. %Note that $B=C$ from dynamical equilibrium
%(Eq. \ref{vshear}). 
We now consider various limits of Eq. \ref{int_rel}.

\subsection{Strictly isothermal gas perfectly coupled to dust}\label{iso_perfect}
When $c_s^2$ is a constant and $\tstop=0$ (so that $\mathcal{C} = 0$), the dusty-gas equations are
exactly equivalent to that for adiabatic hydrodyamics with unit
adiabatic 
index. Although the gas is strictly isothermal, the mixture behaves 
adiabatically because the dust fraction $\tepsilon$ is advected with 
the gas. This is similar to entropy being conserved following an
adiabatic gas when there is no heating or cooling.

In this case, the last two integrals in Eq. \ref{int_rel} vanish. 
Then the condition for axisymmtric \emph{stability}, $\sigma^2>0$, 
is met if the integrand cofficients satisfy $A+D>0$ and $AD-B^2>0$
\citep[][Section 11.6]{ogilvie16}, which becomes 
\begin{align}
  \kappa^2 + \frac{1}{\rhog}\nabla P \cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
  -\frac{1}{\rhog}\frac{\p P}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
\end{align} 
%\begin{align}
%  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
%  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
%    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
%\end{align} 
%\begin{align}
%  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0, \label{dusty_solberg1}    \\
%  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
%  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, \label{dusty_solberg2}
%\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)>0$ is the 
square of the epicylic frequency. Note that $\nabla P/\rhog =
c_s^2\nabla \ln{\rhog}$ for strictly isothermal gas considered here.

@@

On the other hand, Eq. \ref{dusty_solberg2} can be violated in
disks if the dust is vertically well-mixed but radially-stratified,  
such that $\p_z\tepsilon = 0$ but $\p_r\tepsilon\neq0$). In this case the
left-hand-side of Eq. \ref{dusty_solberg2} becomes
\begin{align}
-\left(\frac{1}{\rhog}\frac{\p P}{\p
    z}\right)^2\left(\frac{\p\tepsilon}{\p r}\right)^2<0.
\end{align}
%\begin{align}
%-c_s^4\left(\frac{\p\ln{\rhog}}{\p
%    z}\right)^2\left(\frac{\p\tepsilon}{\p r}\right)^2<0.
%\end{align}
Such an isothermal dusty disk is \emph{unstable} because there is no
effective vertical buoyancy to stabilize vertical motions ($N_z=0$),
which can tap into the free energy associated with  
vertical shear due to the radial gradient in the dust-fraction. 
This situation is identical to the pure gas, adiabtic simulations of \cite{nelson13}
where the gas entropy is vertically uniform but has a radial
gradient. The authors indeed find instability. We give a numerical
example in the dusty context in \S\ref{vert_mixed}.

@@

It is well-known that such correlations can lead to pulsational
instabilities in stars \citep{cox67}. We thus interpret  
dust-drag overstabilities in a similar way, 
adapting from the treatment of stellar 
pulsations by \cite{cox67} and lecture notes by \cite{samadi15} and 
J. Christensen-Dalsgaard\footnote{\url{http://astro.phys.au.dk/$\sim$jcd/oscilnotes/print-chap-full.pdf}}.

>>>>

In general the relative velocity $\bm{v}_\mathrm{d} -
\bm{v}_\mathrm{g}$ obeys a complicated evolutionary equation 
\citep[see, e.g.][]{youdin05a}. % obtained from taking the difference
%between evolutionary equations for $\bm{v}_\mathrm{g}$ and
%$\bm{v}_\mathrm{d}$. 
 However, for tightly-coupled dust
particles with $\tstop\OmK\ll 1 $, where $\OmK$ is the Keplerian orbital
frequency (since we are interested in protoplanetary disks), we can use the 
`terminal velocity approximation' to set %The dust-gas velocity difference 
%Then so that the dust-gas velocity
%difference is 
\begin{align}
  \bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} = \frac{\nabla
    P}{\rhog}\tstop 
\end{align}
\citep{jacquet11}. This equation reflects the well-known effect of 
particle drift towards pressure maximum  \citep{weidenschilling77}.

@@

Under the terminal velocity approximation the dust-gas mixture obeys   
the first-order (in $\tstop$) one-fluid equations: 
\begin{align} 
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
   &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
     \tstop \nabla P \right),\label{dusteq}\\
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi_\mathrm{tot} - \frac{1}{\rho}\nabla  P, \label{momeq}\\ 
  &\frac{D T}{D t} = - \left(\gamma-1\right)T\nabla\cdot\bm{v} +
  \mathcal{H} + \mathcal{H}_\mathrm{eff}  - \Lambda  \label{tempeq} 
\end{align} 
\citep[see ][for a detailed derivation from the two-fluid
equations]{laibe14,price15}, where $D/Dt \equiv \p_t +
\bm{v}\cdot\nabla$ is the Lagrangian 
derivative \emph{following the mixture}'s center-of-mass velocity $\bm{v}$. 
For an ideal gas the pressure is given 
by $P = \mathcal{R}\rhog T/\mu $, where $\mathcal{R}$ is
the gas constant and $\mu$ is the mean molecular weight.

@@

For the mixture's temperature evolution, Eq. \ref{tempeq}, 
$\gamma$ is 
the gas adiabatic index; $\mathcal{H}$ represents heating and $\mathcal{H}_\mathrm{eff}$ 
is an effective source term arising from transforming the gas energy equation (Eq. \ref{real_pressure}) 
from the 
two-fluid to one-fluid variables \citep[see][ for 
details]{laibe14}. We also include a simple model of radiative
cooling, 
\begin{align}
  \Lambda = \frac{T -
    T_\mathrm{ref}}{t_\mathrm{cool}}, \label{cooling} 
\end{align}
which relaxes the mixture back to a prescribed temperature profile   
$T_\mathrm{ref}$ on a timescale of $\tcool$. We will shortly simplify
the problem by considering rapid cooling, $\tcool\to0$.

@@

\subsection{Locally isothermal equation of state}\label{loc_iso_eos}
We consider the limit of short cooling 
times, $\tcool\to 0$, appropriate for the outer parts of an irradiated
protoplanetary disk \citep{chiang97,lin15}. Then the disk temperature
$T = T_\mathrm{ref}$ at all times, and so we may
adopt a locally isothermal equation of state 
\begin{align}\label{eos}
  P = c_s^2(r,z)\rhog = c_s^{2}\left(1 - \tepsilon\right)\rho,   
\end{align}
where $c_s(r,z)= \sqrt{\mathcal{R}T_\mathrm{ref}/\mu}$ is a prescribed
sound-speed profile fixed in time. In most applications we consider vertically 
isothermal disks with \begin{align}\label{power_temp}
  c_s^2(r) \propto r^{q},
\end{align}
where $q$ is the power-law index for the disk temperature. For $q=0$
the disk is strictly isothermal.

Notice Eq. \ref{eos} resembles an ideal gas equation of state but with  
a reduced temperature $\widetilde{T} = 
T_\mathrm{ref}\left(1-\tepsilon\right)$. %sound-speed $c_{s,\mathrm{eff}} = c_s\sqrt{\left(1 -
This reduced temperature decreases with dust-loading. 
%   \tepsilon\right)}$. 
%then the equation of state has the standard
%form $P=c_{s,\mathrm{eff}}^2\rho$. 
Since $\tepsilon$ typically decrease away from the midplane, we expect
vertically isothermal dusty disks to behave as if the temperature
\emph{increased} away from $z=0$.

\subsection{Effective energy equation}\label{energy_analogy}
Although we have deleted the true energy
equation by fixing a locally isothermal
equation of state, we show that the mixture 
nevertheless obeys an effective energy evolution equation. This is
because advection of the dust-fraction, described by
Eq. \ref{dusteq}, can be transformed into an energy-like 
equation.  
%We show that for a prescribed temperature distribution, the mixture
%obeys an evolutionary energy equation, due to the advection of the
%dust-fraction. 
The equation of state, Eq. \ref{eos}, implies 
\begin{align*}
  \tepsilon = 1 - \frac{P}{c_s^2(r,z)\rho}.  
\end{align*}
Then Eq. \ref{dusteq} becomes
\begin{align}
%\frac{DP}{Dt} 
\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  
&= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 -
  \frac{P}{c_s^2\rho}\right)\nabla 
  P\right].
\end{align} 
For comparison with the standard energy equation in hydrodynamics,
re-writing Eq. \ref{tempeq} for a pure ideal gas with $P \propto
\rhog T$, and reverting back to gas velocities, gives  
%For later comparison it is convenient to re-write the temperature
%evolution, Eq. \ref{tempeq}, to that for the pressure: 
\begin{align}
  \frac{\p P}{\p t} + \bm{v}_\mathrm{g}\cdot\nabla P = - \gamma P
  \nabla\cdot \bm{v}_\mathrm{g}  +
  \frac{P}{T}\left(\mathcal{H} -
    \Lambda\right).\label{real_pressure} 
%\frac{\rho \mathcal{R}}{\mu}\frac{T
%  - T_\mathrm{ref}}{t_\mathrm{cool}}, \label{realenergy}
\end{align} 
We can thus interpret Eq. \ref{eff_energy} as the energy equation for
and ideal gas with adiabatic index $\gamma=1$, but now with the
imposed temperature gradient $\nabla c_s^2$ and dust-gas drag $\mathcal{C}$ 
acting as source terms.

@@

The specific entropy of an ideal gas is 
$S~=~C_P\ln{\left(P^{1/\gamma}/\rhog\right)}$, where $C_P$ is the heat 
capacity at constant pressure.

Since we have shown that a locally isothermal dusty gas effectively
has $\gamma=1$ we can define an effective 
entropy for the \emph{mixture} as 
\begin{align}
   \seff \equiv \ln \frac{P}{\rho} = \ln{\left[c_s^2(1-\tepsilon)\right]},  
\end{align} 
where the constant heat capacity has been absorbed into $\seff$. 
%This follows naturally by combining the effective 
%energy Eq. \ref{eff_energy} and Eq. \ref{masseq}. 
Then combining Eq. \ref{eff_energy} and \ref{masseq} gives  
\begin{align*}
  \frac{D \seff}{D t} = \bm{v}\cdot\nabla\ln{c_s^2} +
  \frac{c_s^2}{P}\nabla\cdot\left(\tstop\tepsilon\nabla P\right), 
\end{align*}
which is equivalent to entropy evolution in standard  
hydrodynamics, albeit with source terms. With an effective
entropy defined this way, many of the  
results concerning the stability of (locally) isothermal dusty gas
will have identical form and interpretations to that for pure gas.

@@

%\subsection{Dusty buoyancy forces}
We can now define the vertical buoyancy frequency $N_z$ of the mixture
as    
%$N_z$ of the mixture is given by 
\begin{align}
  N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p \seff}{\p z}
  =c_s^2(r) \frac{\p \ln{\rhog}}{\p z}\frac{\p \tepsilon}{\p z},
%  \quad 
 % N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p \seff}{\p z}.
%&=
  %\frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\frac{\p\ln\rhog}{\p 
%  z}\frac{\p\epsilon}{\p z} \\ &
  %                                =
 % \frac{\epsilon}{\left(1+\epsilon\right)^2}\left(\frac{z}{\Htilde}\right)^2\OmK^2\notag,  
\end{align}
%Typically in a thin, smooth disk we have $|N_r|\ll |N_z|$. 
where the second equality assumes a vertically isothermal disk. 
Protoplanetary disks have $\p_z\rhog<0$, thus stability against vertical 
convection ($N_z^2>0$) requires $\p_z\tepsilon <0 $, i.e. the dust density should drop faster 
than the gas density away from the midplane. 
This is equivalent to 
entropy increasing away from the midplane. A similar expression exists
for the radial buoyancy frequency $N_r$, but $|N_r|\ll |N_z|$ in thin,
smooth disks.

@@

%{\bf
\subsection{Physical disk conditions}
%physical disk conditions underwhich the above approx
%(small/well-coupled dust and isothermal gas) is applicable. smaller
%than mm at few tens of au. give dimensional numbers here
%}
The above correspondence between isothermal dusty gas and adiabatic
pure gas is derived under the terminal velocity approximation, which applies
to short stopping times, $\tstop\OmK\ll 1$; and the locally isothermal
approximation, which applies to short cooling times, $t_\mathrm{cool}\OmK\ll
1$. In typical protoplanetary disk
models such as the Minimum Mass Solar Nebula, $\tstop$ decreases for
smaller particles and/or towards smaller radii \citep{youdin11}. However,
$t_\mathrm{cool}$ is generally only small in the outer disk
\citep{lin15,malygin17}. Combining these results, we estimate the 
correspondence will be applicable to particles less than mm in size at a few
to 10s of AU in protoplanetary disks. However, note that the
isothermal approximation may be relaxed to other fixed equations of
state (see \S\ref{gen_poly}).

@@

For a given distribution of the dust-fraction $\tepsilon$ (or
dust-to-gas ratio $\epsilon$), the 
mass and momentum Eqs. \ref{masseq}---\ref{momeq} admit     
solutions with $\rho(r,z)$ and 
$\bm{v}=r\Omega(r,z)\hat{\bm{\phi}}$ where $\Omega = v_\phi/r$, which satisfy 
\begin{align}
  r\Omega^2 &= \frac{\p \Phi}{\p r} + \frac{1}{\rho}\frac{\p P}{\p
    r},\label{steady_momr}\\
  0 & = \frac{\p\Phi}{\p z} + \frac{1}{\rho}\frac{\p P}{\p z},\label{steady_momz}
%  0 & = \nabla\cdot\left(\tepsilon\tstop\nabla P\right) \label{steady_dust}
\end{align}
with $P=P(\tepsilon,\rho)$ given by the equation of state
(Eq. \ref{eos}). An explicit solution is presented in 
\S\ref{steady_state}, when we analyze the stability of
protoplanetary disks.

@@

Our equilibrium solutions satisfy the hydrostatic constraints of
Eq. \ref{steady_momr}---\ref{steady_momz} but are not in general
steady state solutions to the energy 
equation (\ref{eff_energy}), because the source term $\mathcal{C} \ne0$ for realistic disks.  
Limiting cases with $\mathcal{C} \equiv 0$ include \begin{inparaenum}[1)] 
\item 
  unstratified or 2D, razor-thin disk models with finite dust-gas drag; %(\S\ref{si});   
\item 
  perfectly-coupled dust with $\tstop=0$. %(\S\ref{results}). 
\end{inparaenum} 
The following analyses thus apply strictly to these limiting cases of
$\mathcal{C} = 0$ in equilibrium.  Our analysis is also a good approximation when 
$\tstop$ is sufficiently small such that the evolution of the background disk (e.g. dust-settling) 
occur on much longer timescales than the instability growth timescales.

@@

In Appendix \ref{var_prin} we linearize Eqs. \ref{masseq}, \ref{momeq}, and 
Eq. \ref{eff_energy} to derive the following integral relation: 
\begin{align}
%&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
  \sigma^2\mathcal{I}^2
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \left(\dd v_z \dd v_r^* + \dd v_z^*\dd v_r\right) B +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\Bigl\lvert \nabla\cdot\left(P\dd
  \bm{v}\right)\Bigr\rvert^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV,\label{int_rel}
\end{align} 
%The real integral $\mathcal{I}^2>0$ 
where $^*$ denotes the complex conjugate, 
\begin{align}
  \mathcal{I}^2 \equiv \int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV
\end{align}
is the meridional kinetic energy, 
and coefficients $A,B,D$ can be 
read off Eq. \ref{integral_ex}. %Note that $B=C$ from dynamical equilibrium
%(Eq. \ref{vshear}). 
We now consider various limits of Eq. \ref{int_rel}.

\subsection{Strictly isothermal gas perfectly coupled to dust}\label{iso_perfect}
When $c_s^2$ is a constant and $\tstop=0$ (so that $\mathcal{C} = 0$), the dusty-gas equations are
exactly equivalent to that for adiabatic hydrodynamics with unit
adiabatic 
index. Although the gas is strictly isothermal, the mixture behaves 
adiabatically because the dust fraction $\tepsilon$ is advected with 
the gas. This is similar to entropy conservation following an 
adiabatic gas without heating or cooling.

In this case, the last two integrals in Eq. \ref{int_rel} vanish. 
Then the condition for axisymmtric \emph{stability}, $\sigma^2>0$, 
is met if the integrand coefficients satisfy $A+D>0$ and $AD-B^2>0$
\citep[][Section 11.6]{ogilvie16}, which becomes 
\begin{align}
  \kappa^2 + \frac{1}{\rhog}\nabla P \cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
  -\frac{1}{\rhog}\frac{\p P}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
\end{align} 
%\begin{align}
%  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
%  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
%    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
%\end{align} 
%\begin{align}
%  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0, \label{dusty_solberg1}    \\
%  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
%  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, \label{dusty_solberg2}
%\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)>0$ is the 
square of the epicylic frequency. Note that $\nabla P/\rhog =
c_s^2\nabla \ln{\rhog}$ for strictly isothermal gas considered here.

@@

On the other hand, Eq. \ref{dusty_solberg2} can be violated in
disks if the dust is vertically well-mixed but radially-stratified,  
such that $\p_z\tepsilon = 0$ but $\p_r\tepsilon\neq0$. In this case the
left-hand-side of Eq. \ref{dusty_solberg2} becomes
\begin{align}
-\left(\frac{1}{\rhog}\frac{\p P}{\p
    z}\right)^2\left(\frac{\p\tepsilon}{\p r}\right)^2<0.
\end{align}
%\begin{align}
%-c_s^4\left(\frac{\p\ln{\rhog}}{\p
%    z}\right)^2\left(\frac{\p\tepsilon}{\p r}\right)^2<0.
%\end{align}
Such an isothermal dusty disk is \emph{unstable} because there is no
effective vertical buoyancy to stabilize vertical motions ($N_z=0$),
which can tap into the free energy associated with  
vertical shear due to the radial gradient in the dust-fraction. 
This situation is identical to the pure gas, adiabatic simulations of \cite{nelson13}
where the gas entropy is vertically uniform but has a radial
gradient. The authors indeed find instability. We give a numerical
example in the dusty context in \S\ref{vert_mixed}.

@@

It is well-known that such correlations may lead to pulsational
instabilities in stars \citep{cox67}. We thus interpret  
dust-drag overstabilities in a similar way, 
adapting from the treatment of stellar 
pulsations by \cite{cox67} and lecture notes by \cite{samadi15} and 
J. Christensen-Dalsgaard\footnote{\url{http://astro.phys.au.dk/$\sim$jcd/oscilnotes/print-chap-full.pdf}}.
